Breaking Down the NFL: How Offensive Yards Shape Game Outcomes

### 2024-04-17 paper/paper.qmd
Add files via upload
some quick edits to paper.qmd

When it comes to professional football, there has long been conjecture that a team's performance in games is closely correlated with their offensive yardage total, which includes both passing and running yards. In order to find patterns and insights that could affect team strategy and fan comprehension of the game dynamics, this study conducts a thorough statistical analysis of the complex link between various offensive yard measurements and the probability of winning NFL games. This research attempts to quantify the effect of offensive performance on game outcomes by methodically analyzing play-by-play data for the 2023 NFL season. It provides a thorough summary of how offensive yardage affects a team's victory both throughout the entire game and when focusing only on the statistics for the fourth quarter. Ultimately trying to answer the questions, how do offensive yards impact the outcome of an NFL game, and what is more valuable passing yards or rushing yards[^note]?

[^note]: Rushing yards and running yards are used interchangeably throughout this paper

@@

This study aims to make a substantial contribution to the conversation about sports analytics by offering a strong analytical framework for examining the connection between NFL game-winning and yards. It reveals insights into offensive strategy through in-depth statistical analysis and model-building, providing information for analysts, players and fans. The results of this study show how important data analysis is to sports strategy and how quantitative methods can potentially help improve competitive advantages in professional football.

@@

The foundation of the dataset that was ultimately used in this paper is the play-by-play data from across the 2023 NFL season. This includes an unbelievably dense array of information from every snap over the entirety of the season consisting of 372 variables. These include: seconds left in the half, touchdown probability, expected points added, the probability of taking a safety, the position the snap was taken in, a description of the play, and many, many more. In order to filter this information the play-by-play data was cleaned and new variables were created.

@@

The second logistic regression model, @eq-model2, narrows the focus to the fourth quarter, a critical period where games are often decided. By analyzing the same predictors but confined to the final quarter, this model, $$
\log\left(\frac{p_2}{1-p_2}\right) = \beta_3 + \beta_4 \cdot \text{Q4\_RushingYards} + \beta_5 \cdot \text{Q4\_PassingYards}
$$ {#eq-model2}
, offers insights into the tactical shifts that dominate the endgame. Unlike the overall game model, the fourth-quarter model might reveal a nuanced strategic landscape, such as a negative coefficient for passing yards, indicating the potential pitfalls of a pass-heavy approach under pressure. This focused analysis is instrumental in understanding the dynamics of clutch gameplay, shedding light on the adjustments teams make in response to the scoreboard and the clock.

@@

The choice of logistic regression is prudent in light of the binary structure of the outcome variable (win or loss) and the models' ability to convey the connection between playing strategies (rushing and passing) and winning odds. The interpretability of logistic regression emphasizes its suitability even more; the exponential of the coefficients accurately measures the change in odds that come with every extra yard gained, whether by passing or rushing. This gives you a concise, useful grasp of how various aspects of the game affect your chances of winning.

@@

# 4.1 Graphical Results
```{r, fig.cap= "The Rushing and Passing Yards of Every Game in 2023"}
#| label: fig-1
library(ggplot2)
library(tidyverse)
data <- read.csv("~/nfl_analysis/data/analysis_data/cleaned_data.csv")
ggplot(data, aes(x = RushingYards, y = PassingYards, color = as.factor(Win))) +
  geom_point() +
  labs(
    title = "Rushing Yards vs. Passing Yards by Win/Loss",
    x = "Rushing Yards",
    y = "Passing Yards",
    color = "Outcome"
  ) +
  theme_minimal()
```
@fig-1 shows the rushing yards and passing yards of winning teams over the 2023 NFL season. In examining the graph it is clear that the overall trend, more often than not, is that more rushing yards and more passing yards increase the likelihood of winning. The biggest outlier game, which resulted in the Miami Dolphins Rushing for 350 yards and passing for 376 yards on the Denver Broncos, led to a 70 to 20 victory for the Dolphins. Prior to that game, no team in the 21st century had scored 70 points in a game, this anecdotally shows how being able to run and pass the ball effectively is exceedingly helpful when it comes to scoring points and ultimately winning ball games.

```{r, fig.cap= "Violin Plot of Total Yards by Wins and Losses"}
#| label: fig-2
data <- read.csv("~/nfl_analysis/data/analysis_data/cleaned_data.csv")
ggplot(data, aes(x = as.factor(Win), y = TotalYards, fill = as.factor(Win))) +
  geom_violin() +
  labs(
    title = "Distribution of Total Yards by Win/Loss",
    x = "Outcome (0 = Loss, 1 = Win)",
    y = "Total Yards",
    fill = "Outcome"
  ) +
  theme_minimal()
```
@fig-2 supports this idea, it is a violin plot of the total yards of winning and losing teams (total yards is the sum of passing yards and rushing yards). What it shows is that winning teams never had fewer than 200 total yards, and had more total yardage on average than the teams that lost. It also shows that most winning teams had around 400 total yards while losing teams had just over 350 yards.

```{r, fig.cap= "Average Yards by Wins and Losses"}
#| label: fig-3
data <- read.csv("~/nfl_analysis/data/analysis_data/cleaned_data.csv")
avg_yards_by_outcome <- data |>
  group_by(Win) |>
  summarize(
    AvgRushingYards = mean(RushingYards, na.rm = TRUE),
    AvgPassingYards = mean(PassingYards, na.rm = TRUE)
  ) |>
  pivot_longer(cols = -Win, names_to = "YardType", values_to = "AverageYards")

@@

```{r, fig.cap= "Average Fourth-Quarter Yards by Wins and Losses"}
#| label: fig-4
avg_q4_yards_by_outcome <- read.csv("~/nfl_analysis/data/analysis_data/cleaned_data.csv") %>%
  group_by(Win) %>%
  summarize(
    AvgRushingYards = mean(Q4_RushingYards, na.rm = TRUE),
    AvgPassingYards = mean(Q4_PassingYards, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = -Win, names_to = "YardType", values_to = "AverageYards")

@@

```{r,fig.cap= "The Fourth-Quarter Rushing and Passing Yards of Every Game in 2023"}
#| label: fig-5
q4_final_data_for_modeling <- read.csv("~/nfl_analysis/data/analysis_data/cleaned_data.csv")
# 4th q data Q4_RushingYards,Q4_PassingYards
ggplot(q4_final_data_for_modeling, aes(x = Q4_RushingYards, y = Q4_PassingYards, color = as.factor(Win))) +
  geom_point() +
  labs(
    title = "Fourth-Quarter Rushing vs. Passing Yards by Win/Loss",
    x = "Fourth-Quarter Rushing Yards",
    y = "Fourth-Quarter Passing Yards",
    color = "Outcome"
  ) +
  theme_minimal()
```

```{r, fig.cap= "Distribution of Fourth-Quarter Rushing Yards by Wins and Losses"}
#| label: fig-6
data <- read.csv("~/nfl_analysis/data/analysis_data/cleaned_data.csv")
ggplot(data, aes(x = Q4_RushingYards, fill = as.factor(Win))) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Density of Fourth-Quarter Rushing Yards by Win/Loss",
    x = "Fourth-Quarter Rushing Yards",
    y = "Density",
    fill = "Outcome"
  ) +
  theme_minimal()
```
```{r, fig.cap= "The Distribution of Fourth-Quarter Passing Yards by Wins and Losses"}
#| label: fig-7
data <- read.csv("~/nfl_analysis/data/analysis_data/cleaned_data.csv")
ggplot(data, aes(x = Q4_PassingYards, fill = as.factor(Win))) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Density of Fourth-Quarter Passing Yards by Win/Loss",
    x = "Fourth-Quarter Rushing Yards",
    y = "Density",
    fill = "Outcome"
  ) +
  theme_minimal()
```
Comparing @fig-5 with @fig-1, there is quite a contrast.  Whereas @fig-1 showed that winning teams ran and threw the ball more, @fig-5 shows the winning teams hugging the x-axis (more fourth-quarter rushing yards and fourth-quarter fewer passing yards), with losing teams hugging the y-axis (fewer fourth-quarter rushing yards and more fourth-quarter passing yards).  @fig-6 and @fig-7 show the distributions of fourth-quarter rushing and passing yards by winning and losing teams and both show how pronounced this trend is.

>>>>

When it comes to professional football, there has long been conjecture that a team's performance in games is closely correlated with their offensive yardage total, which includes both passing and running yards. In order to find patterns and insights that could affect team strategy and fan comprehension of the game dynamics, this study conducts a thorough statistical analysis of the complex link between various offensive yard measurements and the probability of winning NFL games. This research attempts to quantify the effect of offensive performance on game outcomes by methodically analyzing play-by-play data for the 2023 NFL season. It provides a thorough summary of how offensive yardage affects a team's victory both throughout the entire game and when focusing only on the statistics for the fourth quarter. Ultimately trying to answer the questions, how do offensive yards impact the outcome of an NFL game, and what is more valuable passing yards or rushing yards[^note]? Prior to examining the data, it was our estimation that in the modern, pass friendly NFL, a focus on throwing the football over the course of the game with a switch to the run game in the fourth-quarter is ultimately the best way for teams to win.

[^note]: Rushing yards and running yards are used interchangeably throughout this paper.

@@

This study aims to make a substantial contribution to the conversation about sports analytics by offering a strong analytical framework for examining the connection between NFL game-winning and yards. It reveals insights into offensive strategy through in-depth statistical analysis and model-building, providing information for analysts, players and fans. The results of this study show how important data analysis is to sports strategy and how quantitative methods can potentially help improve competitive advantages in professional football. This paper is broken up into four sections: the data section which discusses the data used in this paper; the model section explains the models that were used in this paper as well as why they were appropriate; the results section which explains what information was gained through this analysis; and finally the discussion section discusses the findings as a whole.

@@

The foundation of the dataset that was ultimately used in this paper is the play-by-play data from across the 2023 NFL season. This includes an unbelievably dense array of information from every snap over the entirety of the season consisting of 372 variables. These include seconds left in the half, touchdown probability, expected points added, the probability of taking a safety, the position the snap was taken in, a description of the play, and many, many more. In order to filter this information the play-by-play data was cleaned and new variables were created.

The papers data preparation and cleaning process was streamlined, because of how robust the data from NFLVerse [@citenflverse] is. All that needed to be done was the construction of variables (discussed in section 2.2), but otherwise, there was no missing data or impactful decisions that needed to be made during the preparation phase. One thing to note is that the uncleaned dataset is not saved in this repo, this is because the size of the data is too large to store on github so this step was skipped.

@@

The second logistic regression model, @eq-model2, narrows the focus to the fourth quarter, a critical period where games are often decided. By analyzing the same predictors but confined to the final quarter, this model, $$
\log\left(\frac{p_2}{1-p_2}\right) = \beta_3 + \beta_4 \cdot \text{Q4\_RushingYards} + \beta_5 \cdot \text{Q4\_PassingYards}
$$ {#eq-model2}
, offers insights into the tactical shifts that dominate the endgame. Unlike the overall game model, the fourth-quarter model reveals a nuanced strategic landscape, such as a negative coefficient for passing yards, indicating the potential pitfalls of a pass-heavy approach under pressure. This focused analysis is instrumental in understanding the dynamics of clutch gameplay, shedding light on the adjustments teams make in response to the scoreboard and the clock.

@@

The choice of logistic regression is prudent in light of the binary structure of the outcome variable (win or loss) and the models' ability to convey the connection between playing strategies (rushing and passing) and winning odds. The interpretability of logistic regression emphasizes its suitability even more; the exponential of the coefficients accurately measures the change in odds that come with every extra yard gained, whether by passing or rushing. This gives a concise, useful grasp of how various aspects of the game affect a team’s chances of winning.

@@

# 4.1 Graphical Results
```{r, fig.cap= "The Rushing and Passing Yards of Every Game in 2023"}
#| label: fig-1
library(ggplot2)
library(tidyverse)
data <- read.csv(here::here("data/analysis_data/cleaned_data.csv"))
ggplot(data, aes(x = RushingYards, y = PassingYards, color = as.factor(Win))) +
  geom_point() +
  labs(
    title = "Rushing Yards vs. Passing Yards by Win/Loss",
    x = "Rushing Yards",
    y = "Passing Yards",
    color = "Outcome"
  ) +
  theme_minimal()
```
@fig-1 shows the rushing yards and passing yards of winning teams over the 2023 NFL season. In examining the graph it is clear that the overall trend, more often than not, is that more rushing yards and more passing yards increase the likelihood of winning. The biggest outlier game, which resulted in the Miami Dolphins Rushing for 350 yards and passing for 376 yards on the Denver Broncos, led to a 70 to 20 victory for the Dolphins. Prior to that game, no team in the 21st century had scored 70 points in a game, this anecdotally shows how being able to run and pass the ball effectively is exceedingly helpful when it comes to scoring points and ultimately winning ball games.

```{r, fig.cap= "Violin Plot of Total Yards by Wins and Losses"}
#| label: fig-2
data <- read.csv(here::here("data/analysis_data/cleaned_data.csv"))
ggplot(data, aes(x = as.factor(Win), y = TotalYards, fill = as.factor(Win))) +
  geom_violin() +
  labs(
    title = "Distribution of Total Yards by Win/Loss",
    x = "Outcome (0 = Loss, 1 = Win)",
    y = "Total Yards",
    fill = "Outcome"
  ) +
  theme_minimal()
```
@fig-2 supports this idea, it is a violin plot of the total yards of winning and losing teams (total yards is the sum of passing yards and rushing yards). What it shows is that winning teams never had fewer than 200 total yards, and had more total yardage on average than the teams that lost. It also shows that most winning teams had around 400 total yards while losing teams had just over 350 yards.

```{r, fig.cap= "Average Yards by Wins and Losses"}
#| label: fig-3
data <- read.csv(here::here("data/analysis_data/cleaned_data.csv"))
avg_yards_by_outcome <- data |>
  group_by(Win) |>
  summarize(
    AvgRushingYards = mean(RushingYards, na.rm = TRUE),
    AvgPassingYards = mean(PassingYards, na.rm = TRUE)
  ) |>
  pivot_longer(cols = -Win, names_to = "YardType", values_to = "AverageYards")

@@

```{r, fig.cap= "Average Fourth-Quarter Yards by Wins and Losses"}
#| label: fig-4
avg_q4_yards_by_outcome <- read.csv(here::here("data/analysis_data/cleaned_data.csv")) %>%
  group_by(Win) %>%
  summarize(
    AvgRushingYards = mean(Q4_RushingYards, na.rm = TRUE),
    AvgPassingYards = mean(Q4_PassingYards, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = -Win, names_to = "YardType", values_to = "AverageYards")

@@

```{r,fig.cap= "The Fourth-Quarter Rushing and Passing Yards of Every Game in 2023"}
#| label: fig-5
q4_final_data_for_modeling <- read.csv(here::here("data/analysis_data/cleaned_data.csv"))
# 4th q data Q4_RushingYards,Q4_PassingYards
ggplot(q4_final_data_for_modeling, aes(x = Q4_RushingYards, y = Q4_PassingYards, color = as.factor(Win))) +
  geom_point() +
  labs(
    title = "Fourth-Quarter Rushing vs. Passing Yards by Win/Loss",
    x = "Fourth-Quarter Rushing Yards",
    y = "Fourth-Quarter Passing Yards",
    color = "Outcome"
  ) +
  theme_minimal()
```

```{r, fig.cap= "Distribution of Fourth-Quarter Rushing Yards by Wins and Losses"}
#| label: fig-6
data <- read.csv(here::here("data/analysis_data/cleaned_data.csv"))
ggplot(data, aes(x = Q4_RushingYards, fill = as.factor(Win))) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Density of Fourth-Quarter Rushing Yards by Win/Loss",
    x = "Fourth-Quarter Rushing Yards",
    y = "Density",
    fill = "Outcome"
  ) +
  theme_minimal()
```
```{r, fig.cap= "The Distribution of Fourth-Quarter Passing Yards by Wins and Losses"}
#| label: fig-7
data <- read.csv(here::here("data/analysis_data/cleaned_data.csv"))
ggplot(data, aes(x = Q4_PassingYards, fill = as.factor(Win))) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Density of Fourth-Quarter Passing Yards by Win/Loss",
    x = "Fourth-Quarter Rushing Yards",
    y = "Density",
    fill = "Outcome"
  ) +
  theme_minimal()
```
Comparing @fig-5 with @fig-1, there is quite a contrast.  Whereas @fig-1 showed that winning teams ran and threw the ball more, @fig-5 shows the winning teams hugging the x-axis (more fourth-quarter rushing yards and fourth-quarter fewer passing yards), with losing teams hugging the y-axis (fewer fourth-quarter rushing yards and more fourth-quarter passing yards).  @fig-6 and @fig-7 show the distributions of fourth-quarter rushing and passing yards by winning and losing teams and both show how pronounced this trend is.
